Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-26 manuscript/revision2/revision2.tex
revision2: Drop MO from Midwest GR50 mash analysis due to GWAS pop structure issues

Eight GxWeather covariance structures were selected by the greedy
algorithm, one to two per subpopulation: five for green-up date, and
three for flowering date (SI Appendix, Table S2). Of these eight, five
had mass on them in \emph{mash} models of the strong effects
(Figure~\ref{fig-covar} B,D). Three of these five matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while two had all positive or near-zero
covariances.

For green-up date, SNP-associated phenotypic effects covaried with
different weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the \emph{mash} model of Gulf green-up fell on a
covariance matrix constructed using the covariance of day length 14 days
prior to the date of green-up. The covariance matrix for this weather
cue was visually similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A).
\emph{Mash} models of the timing of vegetative growth in the Midwest
subpopulation and Both subpopulations did not include this GxWeather
covariance; the Midwest had no weight on any GxWeather covariance, while
Both subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C) for vegetative growth.

@@

We expected to observe common rank-changing GxE at the level of
individual loci as this is a key theoretical prediction of local
adaptation (26--29). To determine the frequency of rank-changing GxE, we
used the local false sign rate (lfsr), an analogue of the lfdr that
establishes confidence in the effect sign, not the effect's difference
from zero, to determine significance. We required lfsr significance (p
\textless{} 0.05) in both gardens to include effects. This means that
our tests for a sign change between gardens carry an equal statistical
burden to those for effects with the same sign. We separated kinds of
effects at the level of individual loci into SNP effects that differ in
sign between gardens (effects with rank-changing GxE between gardens),
SNP effects that differ in magnitude (effects that are large in one
garden or region and smaller in others), and SNP effects that are
indistinguishable in two gardens (Figure~\ref{fig-effects} A).

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited rank-changing GxE, or a difference in sign
between pairs of Texas and North gardens (Figure~\ref{fig-effects} B; SI
Appendix, Fig S2). 78.7\% of pairwise comparisons between North and
Texas gardens had a difference in sign, while only 28.6\% and 0.2\% of
North-North or Texas-Texas comparisons had a difference in sign,
respectively (Figure~\ref{fig-effects} B). The majority of pairwise
effects for greenup for the Midwest (\textgreater55\%) and Both
(\textgreater85\%) subpopulations were the same sign, and effects most
often differed in magnitude between gardens within and between the
regions (Figure~\ref{fig-effects} B; SI Appendix, Fig S2).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited rank-changing GxE, or a difference in effect sign
within or between regions (Figure~\ref{fig-effects} B). More effects
differed in magnitude between the Texas and North regions than within
these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects} B),
while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering, but a large proportion
of these differed in sign between Texas and North regions (42.7\%) or
within the North region (65.4\%). Finally, in Both subpopulations, less
than 20\% of pairwise effects differed in sign (Figure~\ref{fig-effects}
B). Most differences in sign were between TX1, the southernmost garden,
and all other gardens (SI Appendix, Figure S3). Similarly, more effects
that differed in magnitude included gardens in the Texas region
(52.3-55.9\%), and most effect pairs in the North region were not
distinguishable (91.5\%).

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both
subpopulations revealed many alleles with rank-changing GxE, or sign
changes in their effects between the Texas and North gardens
(Figure~\ref{fig-effects} B). As phenological timings are major
components of plant fitness, this result supports theoretical models
that local adaptation should involve antagonistic pleiotropy at the
level of individual loci (26--29). In addition, it advances local
adaptation research by using more than two field sites and a wide range
of genetic variation to determine that rank-changing GxE for
phenological timings is common in small genomic regions (14, 43, 44).

@@

\includegraphics{images/Figure_2.png}

@@

\includegraphics{images/Figure_3_GxE.png}

@@

\includegraphics{images/Figure_4_QTL_Overlaps.png}

>>>>

Nine GxWeather covariance structures were selected by the greedy
algorithm, one to two per subpopulation: six for green-up date, and
three for flowering date (SI Appendix, Table S2). Of these nine, six had
mass (\textgreater0.1\%) on them in \emph{mash} models of the strong
effects (Figure~\ref{fig-covar} B,D). Five of these six matrices had
negative covariances between gardens in the Texas and North regions
(Figure~\ref{fig-covar} A), while one had all positive or near-zero
covariances.

For green-up date, SNP-associated phenotypic effects covaried with
different weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the \emph{mash} model of Gulf green-up fell on a
covariance matrix constructed using the covariance of day length 14 days
prior to the date of green-up. The covariance matrix for this weather
cue was visually similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A).
\emph{Mash} models of the timing of vegetative growth in the Midwest
subpopulation and Both subpopulations did not include this GxWeather
covariance. The Midwest had 0.15\% weight on a GxWeather covariance
matrix of cumulative GDD for the 28d prior to green-up, a GxWeather
matrix that was visually similar to the Midwest's observed pattern of
phenotypic correlation (Figure~\ref{fig-map} B). Both subpopulations had
non-zero weights on two additional GxWeather covariance types, average
temperature one day prior to green-up, and the day length change in
seconds in the day prior to green-up (Figure~\ref{fig-covar} B). The
average temperature covariance matrix had negative covariances between
Texas and North gardens, though not as strong as the negative phenotypic
correlations seen in Both subpopulations (Figure~\ref{fig-map} B;
Figure~\ref{fig-covar} A). The Gulf subpopulation and Both
subpopulations had substantially more mass on GxWeather covariance
matrices than the Midwest population (Figure~\ref{fig-covar} C) for the
onset of vegetative growth.

@@

We expected to observe common rank-changing GxE at the level of
individual loci as this is a key theoretical prediction of local
adaptation (26--29). To determine the frequency of rank-changing GxE, we
used the local false sign rate (lfsr), an analogue of the lfdr that
establishes confidence in the effect sign, not the effect's difference
from zero, to determine significance. We required lfsr significance (p
\textless{} 0.05) in both gardens to include effects. This means that
our tests for a sign change between gardens carry an equal statistical
burden to those for effects with the same sign. We separated kinds of
effects at the level of individual loci into SNP effects that differ in
sign between gardens (effects with rank-changing GxE between gardens),
SNP effects that differ in magnitude (effects that are large in one
garden or region and smaller in others), and SNP effects that are
indistinguishable in two gardens (similarly large or small in
comparisons between gardens) (Figure~\ref{fig-effects} A).

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited rank-changing GxE, or a difference in sign
between pairs of Texas and North gardens (Figure~\ref{fig-effects} B; SI
Appendix, Fig S2). 78.7\% of pairwise comparisons between North and
Texas gardens had a difference in sign, while only 28.6\% and 0.2\% of
North-North or Texas-Texas comparisons had a difference in sign,
respectively (Figure~\ref{fig-effects} B). The majority of pairwise
effects for greenup for the Midwest (\textgreater73\%; 68-342 effects)
were indistinguishable, or had no GxE. The majority of pairwise effects
for Both (\textgreater85\%) subpopulations were the same sign, and
effects most often differed in magnitude between gardens within and
between the regions; however, these differences in effect sign and
magnitude were mostly between the MO site and other sites
(Figure~\ref{fig-effects} B; SI Appendix, Fig S2).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited rank-changing GxE, or a difference in effect sign
within or between regions (Figure~\ref{fig-effects} B). More effects
differed in magnitude between the Texas and North regions than within
these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects} B),
while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering (0-174 per pairwise
comparison), but a large proportion of these differed in sign between
Texas and North regions (42.7\%) or within the North region (65.4\%).
Finally, in Both subpopulations, less than 20\% of pairwise effects
differed in sign (Figure~\ref{fig-effects} B). Most differences in sign
were between TX1, the southernmost garden, and all other gardens (SI
Appendix, Figure S3). Similarly, more effects that differed in magnitude
included gardens in the Texas region (52.3-55.9\%), and most effect
pairs in the North region were not distinguishable (91.5\%).

@@

Our analysis of the timing of vegetative and reproductive growth
revealed many alleles with rank-changing GxE, or sign changes in their
effects between the Texas and North gardens (Figure~\ref{fig-effects}
B). The Gulf \& Both subpopulations had rank-changing GxE for green-up
date, while the Midwest subpopulation had rank-changing GxE for
flowering date. As phenological timings are major components of plant
fitness, this result supports theoretical models that local adaptation
should involve antagonistic pleiotropy at the level of individual loci
(26--29). In addition, it advances local adaptation research by using
more than two field sites and a wide range of genetic variation to
determine that rank-changing GxE for phenological timings is common in
small genomic regions (14, 43, 44).

@@

\includegraphics{images/Figure_2_six_sites_Midwest.png}

@@

\includegraphics{images/Figure_3_Fraction_GxE_by_Region_six_midwest_sites.png}

@@

\includegraphics{images/Figure_4_QTL_Overlaps_6_Midwest_Gardens.png}
